---
title: 'Assignment 04: Tidy Data & Joins'
author: "Icíar Fernández Boyano"
date: "05/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Full instructions for this assignment can be found **[here](https://stat545.stat.ubc.ca/evaluation/hw04/hw04/).**

### *Set up*

```{r}
library(tidyverse)
library(gapminder)
library(gt) # I used this package to format tables for presentation, as an alternative to knitr
library(wesanderson) # This package contains color palettes from Wes Anderson movies (used in 1.2)
```


### **Exercise 1: Univariate Data Reshaping**

#### **Univariate Option 1**

##### 1. Make a tibble with one row per year, and columns for life expectancy for two or more countries.

*Notes*

+ I used the package **'gt'** to produce a nice-looking table instead of the usual knitr::kable. I used [this](https://gt.rstudio.com/articles/intro-creating-gt-tables.html), and the package is available for download [here](https://github.com/rstudio/gt).

```{r}
gapminder_wide <-
  gapminder %>%
  pivot_wider(id_cols = year,
              names_from = "country",
              values_from = lifeExp) %>% # At this point, we have columns for all countries showing life exp
  select(year, Spain, Canada) %>% # Selecting two countries
  gt() %>%
  tab_header(title = "Life Expectancy over time in Spain and Canada")

gapminder_wide # prints table
```

##### 2. Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

*Notes*

+ I used the package ['wes anderson'](https://github.com/karthik/wesanderson) that has different colour palettes for each of his movies to add colour to this scatterplot. 

```{r}
LifeExp.plot <-
  gapminder_wide %>%
  ggplot(aes(Spain, Canada)) +
  geom_point(color=wes_palette("GrandBudapest1", 12, type = "continuous")) +
  ggtitle("Life Expectancy in Spain and Canada from 1952 to 2007") +
  xlab("Life Expectancy in Spain") +
  ylab("Life Expectancy in Canada") +
  theme_bw() +
  theme(text = element_text(size = 12)) 

LifeExp.plot # Prints plot
```

##### 3. Re-lengthen the data.

```{r}
gapminder_length <- 
  gapminder_wide %>%
  pivot_longer(cols = -year, 
               names_to = "country", 
               values_to = "lifeExp") %>%
  DT::datatable()

gapminder_length # Print
```



### **Exercise 2: Multivariate Data Reshaping**

#### **Multivariate Option 1**

##### 1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

```{r}
# Reshaping data
gapminder_reshape <- 
  gapminder %>%
  pivot_wider(id_cols  = year, 
              names_from = country,
              names_sep = "_", 
              values_from = c("lifeExp", "gdpPercap")) %>% # At this point, data is for all countries
  select(year, lifeExp_Spain, lifeExp_Canada, gdpPercap_Spain, gdpPercap_Canada) 

# Making a nice looking table for presentation

gapminder_reshape %>%
  rename("Life.Spain" = lifeExp_Spain, "Life.Canada" = lifeExp_Canada, "GDP.Spain" = gdpPercap_Spain, "GDP.Canada" = gdpPercap_Canada) %>%
  gt() %>%
  tab_header(title = "Exploring two variables from the Gapminder dataset",
             subtitle = "From 1952 to 2007") %>%
  tab_spanner(label = "Life Expectancy",
              columns = vars(Life.Spain, Life.Canada)) %>%
  tab_spanner(label = "GDP Per Capita",
              columns = vars(GDP.Spain, GDP.Canada))

```


##### 2. Re-lengthen the data.

```{r}
gapminder_reshape %>%
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_") %>%
  DT::datatable()
```

### **Exercise 3: Table Joins**

Read in the made-up wedding guestlist and email addresses using the following lines (go ahead and copy-paste these):

```{r, echo=FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

##### **3.1: For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.**

###### *Notes*

+ I found [this thread](https://stackoverflow.com/questions/13773770/split-comma-separated-strings-in-a-column-into-separate-rows) particularly useful to solve this question, as I got a bit stuck with organising the email tibble.

```{r}
# Splitting the comma-divided rows into individual columns
email <- 
  as_tibble(email) %>%
  mutate(name = strsplit(as.character(guest), ",")) %>%
  unnest(name)

# Joining the tibbles
guest %>%
  left_join(email, by = "name") %>% # Merge by name variable
  select(name, email) %>% # Only keep the guest names & emails 
  DT::datatable() # Tibble presentation
```


##### **3.2: Who do we have emails for, yet are not on the guestlist?**

```{r}
email %>%
  anti_join(guest, by = "name") %>%
  knitr::kable()
```


##### **3.3: Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).**

###### *Notes*

+ Guests Atlanta and Denzel Connolly, Chanelle Shah, Cosmo Dunkley, and Cai Mcdaniel are on the guest list but do not have emails - we can see this with full_join.

```{r}
full_join(guest, email, by = "name") %>%
  DT::datatable()
```

+ If we want to only include those guests that have email addresses from both tibbles (email and guest), we can omit those rows that have NA values for the email column in the full_join tibble.

```{r}
full_join(guest, email, by = "name") %>%
  select(name, email) %>%
  na.omit() %>%
  knitr::kable()
```




